---
title: "exercises"
format: html
editor: visual
---

#Notes: - use command, shift, m to pipe - pipe in between each new line - pipe is also %\>% - control, option, i to create new code chunk

#Install

```{r}
library(dplyr)
```


#Questions

```{r}
#1
tree_dat <- read.csv("~/github/Lab_02/data/doi_10_5061_dryad_18pm5__v20170130/Itter_et_al_EAP16-0589.R1/tree_dat.csv")
glimpse(tree_dat)
```

I see that there are 131,386 rows and 8 columns.

```{r}
#2
tree_dat %>%
  filter(standID == 1) %>%
  tally()
```

There have been 979 records made in stand 1.

```{r}
#3
tree_dat %>%
  filter(species %in% c("ABBA","PIST")) %>%
  count(species)
```

Abies balsamea: 13,033 Pinus strobus: 4,188

```{r}
#4
last_year <- max(tree_dat$year, na.rm = TRUE)

tree_dat %>%
  filter(year == last_year, age > 200) %>%
  count()
```

There are 7 trees older than 200 years old in the last year of the dataset.

```{r}
#5
tree_dat %>%
  slice_max(order_by = age, n=1)
```

The oldest tree in the dataset is Pinus resinosa, it's tree ID is 24.

```{r}
#6
tree_dat %>%
  filter(year == "2001") %>%
  slice_max(order_by = age, n=5) 
  
```

The five oldest trees recorded in 2001 are (top 2): Pinus resinosa and (last 3): Fraxinus nigra. The tree IDs are 24, 25, 1595, 1598, and 1712.

```{r}
#7
tree_dat %>%
  filter(year == "2002") %>%
  slice_sample(prop = .3)
```

There are 687 trees in a 30% sample of those recorded in 2002.

```{r}
#8
tree_dat %>% 
  filter(standID == 5, year == "2007") %>%
  arrange(desc(rad_ib)) %>%
  slice_head(n = 3)
```

The tree IDs are 128, 157, and 135

```{r}
#9
tree_dat %>% 
  select(treeID, standID, year, rad_ib) %>% 
  filter(standID == 3, year == "2007") %>% 
  slice_min(order_by = rad_ib, n = 3)
```

The three smallest trees measure that year had the tree IDs of 50, 56, and 36.'

```{r}
#10
tree_dat %>% 
  select(-stand)
glimpse(tree_dat)
```
The stand column was removed.


```{r}
# 11
tree_dat %>% 
  select(matches("ID"))
glimpse(tree_dat)
```
The columns treeID and standID were shown.

```{r}
# 12
tree_dat %>% 
  select(matches("ID|stand"))
glimpse(tree_dat)
```
The columns treeID, standID, and stand were shown.

```{r}
# 13
tree_dat <- tree_dat %>% 
  mutate(
    rad_ib_mm = rad_ib,
    inc_mm = inc
  )
glimpse(tree_dat)
```
This renaming was permanent.

```{r}
# 14
tree_dat <- tree_dat %>% 
  mutate(
    rad_ib_cm = ifelse(rad_ib_mm > 10, rad_ib_mm / 10, rad_ib_mm),
    DBH_cm = 2 * rad_ib_cm,
    BA_m2 = 0.00007854 * (DBH_cm^2)
  )

glimpse(tree_dat)

mean_BA_m2_POTR_2007 <- tree_dat %>%
  filter(species == "POTR", year == 2007) %>% 
  summarize(mean_BA = mean(BA_m2, na.rm = TRUE))
```
The mean BA_m2 of the species POTR in 2007 was 0.03696619. 

```{r}
# 15
tree_dat <- tree_dat %>% 
  mutate(established = if_else(age > 5, TRUE, FALSE))

established_count <- tree_dat %>% 
  count(established)
```
There are 8883 trees younger or equal to the age of 5. There are 122503 trees older than the age of 5.

```{r}
# 16
tree_dat <- tree_dat %>% 
  mutate(
    DBH_class = case_when(
      DBH_cm >= 0 & DBH_cm < 2.5  ~ "seedling",
      DBH_cm >= 2.5 & DBH_cm < 10 ~ "sapling",
      DBH_cm >= 10 & DBH_cm < 30  ~ "pole",
      DBH_cm >= 30               ~ "sawlog",
      TRUE ~ NA_character_
    )
  )

DBH_class_counts_2007 <- tree_dat %>% 
  filter(year == 2007) %>% 
  count(DBH_class)
```

There are 0 seedlings, 252 saplings, 1963 poles, and 76 sawlogs.

```{r}
# 17
DBH_stats_2007 <- tree_dat %>% 
  filter(year == 2007) %>%
  summarize(
    mean_DBH = mean(DBH_cm, na.rm = TRUE),
    sd_DBH = sd(DBH_cm, na.rm = TRUE)
  )
```

The mean of the DBH is 16.09351. The standard deviation of the DBH is 6.138643.

```{r}
# 18
mean_age_per_species_2003 <- tree_dat %>% 
  filter(year == 2003) %>% 
  group_by(species) %>% 
  summarize(mean_age = mean(age, na.rm = TRUE)) %>% 
  arrange(desc(mean_age)) %>% 
  slice_head(n = 3)
```

In 2003, the mean age of Thuja occidentalis was 126.63830. The mean age of Fraxinus nigra was 83.08333. The mean age of Pinus strobus was 73.28571.

```{r}
# 19
years <- tree_dat %>% 
  summarize(
    unique_years = n_distinct(year),
    first_year = min(year, na.rm = TRUE),
    last_year = max(year, na.rm = TRUE)
  )
```

There were 111 unique years. 1897 was the first year ever recorded. 2007 was the last year ever recorded.

```{r}
# 20
stands_with_most_years <- tree_dat %>% 
  group_by(stand) %>% 
  summarise(unique_years = n_distinct(year)) %>% 
  filter(unique_years == max(unique_years))
```

The stands that are tied with the largest temporal record are A1, D1, D2, D3, and F1.

```{r}
tree_dat <- tree_dat %>% 
  mutate(DBH_cm = rad_ib * 2 / 10)

annual_DBH_growth <- tree_dat %>% 
  arrange(treeID, year) %>% 
  group_by(treeID) %>% 
  mutate(annual_growth = DBH_cm - lag(DBH_cm)) %>% 
  filter(!is.na(annual_growth))

tree_growth_summary <- annual_DBH_growth %>% 
  group_by(treeID) %>% 
  summarise(total_years = n(), avg_growth = mean(annual_growth, na.rm = TRUE)) %>% 
  filter(total_years >= 10)

tree_growth_species <- tree_growth_summary %>% 
  left_join(tree_dat %>% select(treeID, species), by = "treeID") %>% 
  distinct()

species_growth <- tree_growth_species %>% 
  group_by(species) %>% 
  summarise(mean_growth = mean(avg_growth, na.rm = TRUE),
            sd_growth = sd(avg_growth, na.rm = TRUE)) %>% 
  arrange(desc(mean_growth))

fastest_species <- species_growth %>% 
  slice_head(n = 3)

slowest_species <- species_growth %>% 
  slice_tail(n = 3)
```

The three fastest growing species are (1) Pinus resinosa, (2) Pinus banksiana, and (3) Populus tremuloides. The three slowest growing species are (1) Quercus rubra, (2) Larix laricina, and (3) Thuja occidentalis. 
